Gaia's constraints on the $\sim$$150\,M_\odot$ runaway star candidate VFTS682

### 2018-06-09 VFTS682_APJL.tex
minor
new fig.1 with thicker hatches and table
added figure on angles and text: the fig needs to be redone with errors properly

\begin{table}
  \begin{center}
    \caption{Astrometric parameters for VFTS682. }
    \begin{tabular}{l|c|c}
      \hline
      \hline
      Parameter & Value & Source\\
      \hline
      RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
                        & \multirow{2}{*}{\cite{evans:11}}\\[5pt]
      DEC \hfill [degree] & -69.07 % $\pm$  0.05
                        & \\[5pt]
      \hline
      $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.10$\pm$0.07 & \multirow{2}{*}{\emph{Gaia} DR2}\\[5pt]
      $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.09 & \\[5pt]
      \hline
      $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
      $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                        0.2$\pm$0.1 &
      \\[5pt]
      \hline
      $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
      \hline
    \end{tabular}
    \tablecomments
    {The error on the RA and DEC positions, are of order
      $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$ in \emph{Gaia}
      DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

@@

Assuming that the best estimate of the peculiar velocity of VFTS682
are reliable, we now address the question of its likely origin.

Figure~\ref{fig:mu_dist} shows the distribution in proper motion relative to R136
of OB type and Wolf-Rayet stars included both in the VFTS survey and
\emph{Gaia DR2} with reliable astrometric solutions. VFTS682 is not an
extreme outlier compared to the two O-type runaways (re-)identified by
\cite{lennon:18}. VFTS682 was expected to be a ``slow runaway'' by \citet{bestenlehner:11} in
the dynamical ejection scenario. The green shade in
\Figref{fig:mu_dist} shows the range of relative proper motions within
the uncertainties on the apparent age of the star, assuming it was
ejected from R136 very early in its life. The proper motion
measurements from \emph{Gaia} DR2 are consistent with the hypothesis
of dynamical ejection.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/dist_mu_region.pdf}
  \caption{Distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an extreme outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The dark blue histogram contains 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histogram contains 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.}
  \label{fig:mu_dist}
\end{figure}

@@

Although the large uncertainties on the relative proper motion
components results in a wide range of possible directions,we argue that the most likely origin of the star is R136.

The kinematic age of this star, assuming it originates from the
cluster, is

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu^\mathrm{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\,\mathrm{mas\ yr^{-1}}} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of R136 \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.

The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age
of $1.0\pm 0.2$\,Myr \citep{schneider:18}. % , which
% corroborates the idea that the star is the result of a dynamical
% ejection very soon in the cluster evolution.
Note that the
present-day surface helium abundance
\citep[$Y\simeq0.5$,][]{bestenlehner:11, rubio-diez:17} puts a lower
limit on the age of the star of $\sim$0.9\,Myr, corresponding to the time needed to
synthesize this amount of helium in the models from \cite{kohler:15}.
% \todo{discuss position angle}

@@

\citet{lennon:18} carried out a study similar to ours on the fast
moving O-type stars
in the region, and found two massive runaway stars
($\sim$$90\,M_\odot$) in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS 72) is less clear given its direction
of motion. The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 (see \Secref{sec:r136_origin}) is smaller than the
corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic
age of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

>>>>

\begin{table}
  \begin{center}
    \caption{Observed and derived parameters for VFTS682. We also list
    the average proper motion components of R136 derived by
    \cite{lennon:18} for reference.}
    \begin{tabular}{l|c|c}
      \hline
      \hline
      Parameter & Value & Source\\
      \hline
      RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
                        & \multirow{2}{*}{\cite{evans:11}}\\[5pt]
      DEC \hfill [degree] & -69.07 % $\pm$  0.05
                        & \\[5pt]
      \hline
      $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.10$\pm$0.07 & \multirow{2}{*}{\emph{Gaia} DR2}\\[5pt]
      $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.09 & \\[5pt]
      \hline
      $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
      $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                        0.2$\pm$0.1 &
      \\[5pt]
      \hline
      $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
      \hline
      present day $M$ \hfill $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ &
                                                     \multirow{3}{*}{\cite{schneider:18}}\\
      $M_\mathrm{ZAMS}$\hfill $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ &
      \\
      age \hfill [Myr] & $1.0\pm0.2$ & \\
      \hline
      $\langle\mu_\mathrm{RA}\rangle$\hfill[$\mathrm{mas\ yr^{-1}}$] & $1.74\pm0.01$
                        &  \multirow{2}{*}{\cite{lennon:18}}\\[5pt]
      $\langle\mu_\mathrm{DEC}\rangle$\hfill[$\mathrm{mas\ yr^{-1}}$]
                & $0.70\pm0.02$ & \\
      \hline

    \end{tabular}
    \tablecomments
    {The error on the RA and DEC positions, are of order
      $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$ in \emph{Gaia}
      DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

% \begin{table}
%   \begin{center}
%     \caption{Astrometric parameters for VFTS682. }
%     \begin{tabular}{l|c|c}
%       \hline
%       \hline
%       Parameter & Value & Source\\
%       \hline
%       RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
%                         & \multirow{2}{*}{\cite{evans:11}}\\[5pt]
%       DEC \hfill [degree] & -69.07 % $\pm$  0.05
%                         & \\[5pt]
%       \hline
%       $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.10$\pm$0.07 & \multirow{2}{*}{\emph{Gaia} DR2}\\[5pt]
%       $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.09 & \\[5pt]
%       \hline
%       $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
%       $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
%                                                                         0.2$\pm$0.1 &
%       \\[5pt]
%       \hline
%       $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
%       \hline
%     \end{tabular}
%     \tablecomments
%     {The error on the RA and DEC positions, are of order
%       $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$ in \emph{Gaia}
%       DR2. %REPETITION: %  We adopt the peculiar
%       % radial velocity calculated using the  NV $\lambda4944$ line, which is
%       % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
%     }
%   \end{center}
%   \label{tab:vfts682}
% \end{table}

@@

% Assuming that the best estimate of the peculiar velocity of VFTS682
% are reliable, we now address the question of its likely origin.

Figure~\ref{fig:mu_dist} shows the distribution in proper motion relative to R136
of OB-type and Wolf-Rayet stars included both in the VFTS survey and
\emph{Gaia DR2} with reliable astrometric solutions. The bottom panel
shows the distribution in angles $\theta$ between the relative proper motion
direction and the direction from the core of R136 to the star. We
emphasize that our subset of stars with reliable proper motion
measurements is biased towards the fast moving objects, which results in a
distribution of angles mildly peaked at small angles.

VFTS682 is not an outlier in relative proper motion%  compared
% to the two O-type runaways (re-)identified by \cite{lennon:18} (see
% also below)
. However, the star was expected to be a ``slow runaway'' by \citet{bestenlehner:11} in
the dynamical ejection scenario. The green shade in
\Figref{fig:mu_dist} shows the range of relative proper motions
required to reach the present day location within
the uncertainties on the apparent age of the star, assuming it was
ejected from R136 very early in its life.

Because of the large
error bars, the angle between the relative proper motion is not very
constraining, but it is suggestive that the best value is close to zero.
Therefore, the relative proper motion from \emph{Gaia} DR2 are consistent with the hypothesis
of dynamical ejection.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/dist_mu_region.pdf}\\
  \includegraphics[width=0.5\textwidth]{figures/angle_dist}
  \caption{Top panel: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. Bottom panel: \todo{redo with
      errorbars} distribution of
    angles between the relative proper motion direction and the radial
    direction to the star. The error bars for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$. }
  \label{fig:mu_dist}
\end{figure}

@@

% Although the large uncertainties on the relative proper motion
% components results in a wide range of possible directions,we argue that the most likely origin of the star is R136.

The kinematic age of this star, assuming it originates from R136, is

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu^\mathrm{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\,\mathrm{mas\ yr^{-1}}} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age
of $1.0\pm 0.2$\,Myr \citep{schneider:18}. % , which
% corroborates the idea that the star is the result of a dynamical
% ejection very soon in the cluster evolution.

% the thing below is not relevant if it is a merger, better cut out
% for now, since we are too long anyway

%Note that the
% present-day surface helium abundance
% \citep[$Y\simeq0.5$,][]{bestenlehner:11, rubio-diez:17} puts a lower
% limit on the age of the star of $\sim$0.9\,Myr, corresponding to the time needed to
% synthesize this amount of helium in the models from \cite{kohler:15}.

@@

\todo{maybe move around fig. 1}
\citet{lennon:18} carried out a study similar to ours on the fast
moving O-type stars
in the region, and found two massive runaway stars
($\sim$$90\,M_\odot$) in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS 72) is less clear given its direction
of motion. The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 (see \Secref{sec:r136_origin}) is smaller than the
corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic
age of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.
